\sys: Controllable Program Regeneration

### 2025-05-22 paper/draft.tex
fix template
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>
improve example
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

\documentclass[sigplan,anonymous]{acmart}

\acmSubmissionID{PAPER ID}
\renewcommand\footnotetextcopyrightpermission[1]{}
\settopmatter{printfolios=true,printacmref=false}
\acmConference[Conference acronym 'XX]{Make sure to enter the correct
  conference title from your rights confirmation email}{June 03--05,
  2018}{Woodstock, NY}

\usepackage{setspace}
\usepackage{enumerate}
\usepackage{algorithm2e}
\usepackage{algpseudocode}
\usepackage{graphics}
\usepackage{xparse} 
\usepackage{xspace}
\usepackage{multirow}
\usepackage{csvsimple}
\usepackage{balance}
\usepackage{minted}
\usepackage{xcolor}
\usepackage{nicefrac}
\usepackage{siunitx}
\usepackage{array,framed}
\usepackage{booktabs}
\usepackage{
  color,
  float,
  epsfig,
  wrapfig,
  graphics,
  graphicx,
  subcaption
}
\usepackage{adjustbox}
\usepackage{csquotes}

\captionsetup{font=footnotesize,labelfont={bf, footnotesize}, belowskip=2pt}

\newcommand{\TODO}[1]{\hl{\textbf{TODO:} #1}\xspace}
\newcommand{\sys}{{\scshape TOOL}\xspace}
\newcommand{\rf}[1]{\ref{#1}}
\newcommand{\sx}[1]{(\S\ref{#1})}
\newcommand{\cf}[1]{(\emph{Cf}.\S\ref{#1})}
\newcommand{\se}[1]{\S\ref{#1}}
\newcommand{\fg}[1]{Fig.~\ref{#1}}
\newcommand{\heading}[1]{\vspace{2pt}\noindent\textbf{\emph{#1}}:\enspace}
\newcommand{\ttt}[1]{\mintinline[fontsize=\normalsize, breaklines, breakafter=-]{js}{#1}}

@@

\title{Submission ID}

\begin{abstract}
Test Abstract
\end{abstract}

@@

>>>>

\documentclass[sigplan]{acmart}

% \acmSubmissionID{PAPER ID}
\renewcommand\footnotetextcopyrightpermission[1]{}
\settopmatter{printfolios=true,printacmref=false}
\acmConference[Conference acronym 'XX]{Make sure to enter the correct
  conference title from your rights confirmation email}{June 03--05,
  2018}{Woodstock, NY}

\usepackage{setspace}
\usepackage{enumerate}
\usepackage{algorithm2e}
\usepackage{algpseudocode}
\usepackage{graphics}
\usepackage{xparse} 
\usepackage{xspace}
\usepackage{multirow}
\usepackage{csvsimple}
\usepackage{balance}
\usepackage{minted}
\usepackage{xcolor}
\usepackage{nicefrac}
\usepackage{siunitx}
\usepackage{array,framed}
\usepackage{booktabs}
\usepackage{
  color,
  float,
  epsfig,
  wrapfig,
  graphics,
  graphicx,
  subcaption
}
\usepackage{adjustbox}
\usepackage{csquotes}
\usepackage{cleveref}

\captionsetup{font=footnotesize,labelfont={bf, footnotesize}, belowskip=2pt}
\def\eg{{\em e.g.}, }
\def\ie{{\em i.e.}, }
\def\etc{{\em etc.}\xspace}
\def\vs{{\em vs.}\xspace}

\newcommand{\TODO}[1]{\hl{\textbf{TODO:} #1}\xspace}
\newcommand{\sys}{{\scshape Sys}\xspace}
\newcommand{\rf}[1]{\ref{#1}}
\newcommand{\sx}[1]{(\S\ref{#1})}
\newcommand{\cf}[1]{(\emph{Cf}.\S\ref{#1})}
\newcommand{\se}[1]{\S\ref{#1}}
\newcommand{\fg}[1]{Fig.~\ref{#1}}
\newcommand{\heading}[1]{\vspace{2pt}\noindent\textbf{\emph{#1}}:\enspace}
\newcommand{\ttt}[1]{\mintinline[fontsize=\normalsize, breaklines, breakafter=-, bgcolor=white]{javascript}{#1}}
\newcommand{\xxx}{\colorbox{red}{xxx}\xspace}

@@

\title{\sys: Controllable Program Regeneration}
\author{Evangelos Lamprou}

% \sys is a framework for regenerating programs.
% \sys accepts a variety of wanted program properties that should be preserved after regeneration.
% \sys also takes a set of guiding properties that will be given as input to the regeneration component.
% Then, \sys will regenerate the program to satisfy the wanted properties while preserving the guiding properties.
% \sys can be used for source to source transformation, program repair, (original code can be: insecure, slow, proprietary, wrong, bad context, different language, non-idiomatic).
% \sys uses a plugin architecture that allows extending it with program property extractors (IO pairs, CFG, AST, execution traces, execution time, etc.).
% The regeneration component can be extended with different regeneration algorithms (e.g., genetic programming, symbolic execution, LLM, etc.).

% Use cases for something like this include:
% - program repair
% - turning insecure code into secure code by removing side-effects
% - transforming a program in language A to language B
% - turning a program into a more idiomatic version of the same language
% - having a program use a different API
% - have a program be more amendable to parallelization
% - have a program be more amendable to further analysis or transformation

\begin{abstract}
\end{abstract}

@@

\section{Example}
\sys enables users to specify transformation goals declaratively.
These goals are expressed as logical constraints over properties of the
regenerated program, such as input-output equivalence, side-effect elimination,
target language, modular structure, and others.

Given a program and a query over desired properties, \sys selects a minimal set
of properties to extract from the original program, then attempts to synthesize
a new program that satisfies the query.
This process is cast as a
minimum-satisfiability (MinSAT) problem, optimizing for the smallest extraction
effort necessary to enable transformation.
If the query is unsatisfiable under
the available constraints or transformations, \sys issues an error
message, showing the conflict.

This section goes through an example applying \sys
to five distinct program transformation tasks, 
targeting security, language translation, compatibility,
idiomaticity, and modularity.

\heading{A compromised JavaScript library}
The \texttt{flatmap-stream} library was implicated in a high-profile
supply-chain attack that exfiltrated Bitcoin credentials.
Although it preserved
client-observable behavior, the library accessed the file system, network, and
global state under specific conditions.
Program regeneration can be used to automatically remove possible malicious payloads
while preserving the original program's observable behavior,
especially in this scenario, where the attack is highly obfuscated and 
activates under very precise conditions.
~\cite{harp:ccs:2021}.
\begin{minted}[frame=lines]{js}
module.exports = function (stream, fn) {
  if (process.env.PRODUCTION) {
    const priv = copay.getKeys()?.[0]?.priv;
    if (priv && wallet.balance > 0) {
      http.request({ hostname: 'evil.com' })
          .end({ priv });
    }}
  return flatten(steam).apply(fn);
};
\end{minted}

\begin{wrapfigure}[3]{r}{.5\columnwidth}
\vspace{-10pt}
\begin{minted}{prolog}
equal(io(P), io(P_)).
pure(P_).
\end{minted}
  % \begin{minted}{shell}
% $ sys -i flatmap-stream.js -o flatmap-stream.js -q query.pl
  % \end{minted}
\end{wrapfigure}
A program written in the \sys DSL to express this intent is shown on the right.
This query requires the regenerated program \ttt{P_} to exhibit the same I/O behavior
but forbids observable side effects.
\sys first arrives at a minimal set of properties to extract from the
original program, which will preserve the requested behavior, while increasing confidence that unwanted properties are retained.
The \sys logic engine produces the set $\{\texttt{funcs(F, P), io(F), sig(F), pure(P)}\}$, where
\texttt{funcs(F, P)} extracts the set of functions defined in the original program
using a language-aware component that either parses the program source or extracts
symbols from the binary.
original program, \texttt{io(F)} extracts the I/O behavior of each function
by providing its source code to an LLM instance and asking it to generate a set of
inputs; corresponding outputs are then generated by executing the original program.
\texttt{sig(F)} extracts each function's signature.
The, \texttt{pure(P)} property means that \sys will confirm after regeneration that the original program is (apparently) pure
using static analysis.

After \xxx seconds, \sys generates the following program:
\begin{minted}[frame=lines]{js}
module.exports = function(stream, fn) {
  return stream.map(fn); };
\end{minted}

\heading{Language translation}
\sys takes advantage of the I/O intermediate representation to
\texttt{flatmap-stream} into Haskell, while preserving its observable
semantics.
\begin{wrapfigure}[3]{r}{.5\columnwidth}
\vspace{-10pt}
\begin{minted}{prolog}
equal(io(P), io(P_)).
language(P_, haskell).
\end{minted}
\end{wrapfigure}
The motivation may be integration into a Haskell-based pipeline or
enabling formal verification.
This transformation goal is expressed as the query shown on the right.
Given this query, \sys again extracts I/O examples from the original program,
then it transforms each eaxmple into an equivalent one in Haskell (\eg 
the pair $\langle\texttt{([1,[2,3]], (x)=>x+1)}\to\texttt{[2,3,4]}\rangle$ 
becomes $\langle(\texttt{[1, [2,3]], Js("(x)=>x+1"))}\to\texttt{[2,3,4]}\rangle$).
Then, it provides the transformed examples to a new LLM instance, prompting it
to generate a Haskell program that satisfies the same I/O behavior.
\begin{minted}[]{haskell}
flatMap :: (a -> b) -> [Either a [a]] -> [b]
flatMap fn stream = concatMap fn stream
\end{minted}
The I/O examples are then again used to confirm the correctness of the
regenerated program.

\heading{An un-idiomatic C program}
We begin with the classic fast inverse square root routine from the Quake III
engine. 
The original implementation exploits type punning to manipulate IEEE
floats at the bit level---an optimization that relies on undefined behavior:

\begin{minted}[]{c}
float Q_rsqrt(float number) {
 long i; float x2, y;
 const float threehalfs = 1.5F;
 x2 = number * 0.5F; y  = number;
 i  = * ( long * ) &y; // evil bit level hack
 i  = 0x5f3759df - ( i >> 1 );
 y  = * ( float * ) &i;
 y  = y * ( threehalfs - ( x2 * y * y ) );
 return y;
}
\end{minted}

The user provides the following query to \sys:
\begin{wrapfigure}[3]{r}{.5\columnwidth}
  \vspace{-10pt}
  \begin{minted}{prolog}
    equal(io(P), io(P_)).
    #min len(P_).
  \end{minted}
\end{wrapfigure}
\sys extracts only the I/O trace and basic arithmetic operations from the
original implementation, and regenerates a new version that avoids undefined
behavior.
As a coarse metric of the produced code's idiomaticity, \sys uses the
length of the generated program, which is minimized in the query.
When instructed to optimize an aspect of the program, \sys will re-prompt
the model to regenerate the program until either no further improvements are made after two consecutive iterations or
the difference between two consecutive generations is below a user-defined threshold.
After \xxx seconds, and \xxx iterations, \sys produces the following program:
\begin{minted}[fontsize=\small]{c}
#include <math.h>
float Q_rsqrt(float number) {
    return 1.0f / sqrtf(number);
}
\end{minted}
The regenerated program is now idiomatic C code, lending itself 
to higher portability, and easier analysis either by the compiler 
or downstream tools.

\heading{A rigid music database application}
Consider a JavaScript application that retrieves and displays a user's music
collection from an SQLite database:

\begin{minted}[fontsize=\small]{js}
// Original version: synchronous, monolithic
function getAlbumsByArtist(artist) {
  const db = new Database("music.db");
  const rows = db.prepare("SELECT album FROM songs WHERE artist = ?").all(artist);
  return rows.map(row => row.album);
}
\end{minted}
We use \sys to produce a modular, asynchronous version compatible with modern
JS environments and better separation of concerns:

\begin{minted}[fontsize=\small]{js}
// Regenerated: modular and async
async function connectDB() {
  const sqlite = require('sqlite');
  return sqlite.open({ filename: 'music.db', driver: sqlite.Database });
}

async function getAlbumsByArtist(db, artist) {
  const rows = await db.all("SELECT album FROM songs WHERE artist = ?", artist);
  return rows.map(row => row.album);
}
\end{minted}
The regenerated code is now more modular.

% Snippets:

% Q_rsqrt: security, idiomaticity, language-translation
% Music Collection DB Access: compatibility, modularity

\section{Evaluation}

% benchmark summary table

\begin{table}[h]
\centering
\caption{Summary of benchmarks used in the evaluation.}
\begin{tabular}{llll}
\toprule
Benchmark                          & Description                 & Size & Source \\
\midrule
Rosetta Code                       & Programming tasks           &      & \\
npm utilities                      & Utilities from npm          &      & \\
Python utilities                   & Utilities from PyPi         &      & \\
SSCA                               & Sneaky supply-chain attacks &      & \\
IOCCC                              & Obfuscated C code           &      & \\
Microbenchmarks                    & Custom microbenchmarks      &      & \\
\hspace{.5em} \ttt{flatmap-stream} &                             &      & \\
\hspace{.5em} \ttt{Q_rsqrt}        &                             &      & \\
\hspace{.5em} \textsf{MusicDB}     &                             &      & \\
\bottomrule
\end{tabular}
\label{tab:benchmarks}
\end{table}
